Data Exploration: IUCN and AquaMaps

### 2015-09-11 data_explore.Rmd
added aquamaps threshold exploration

library(ggplot2)
library(maps)
library(readr)
library(raster)
library(tidyr)
library(dplyr)

@@

  p1 = perc_c %>% dplyr::select(percent)
  paste0(round(p1[1, ], 2), '% of the total area in this map is covered by both IUCN and AquaMaps')
  paste0(round(p1[2, ], 2), '% of the total area in this map is covered just by AquaMaps')
  paste0(round(p1[3, ], 2), '% of the total area in this map is covered just by IUCN')
```

@@

This plot shows that the majority of species range maps have only ~30% of their full map covered by both maps, and almost no species have more than 50% of their two ranges (IUCN and AquaMaps) overlapping.

@@

all<-ggplot(df,aes(percent,fill=perc.type))+
    geom_histogram(binwidth=1,alpha=.3)+
    scale_fill_discrete(name="Legend",labels=c("Overlap", "AquaMaps only", "IUCN only"))+
    labs(x='Percent of total range', y='Number of Species')

@@

sp <- ggplot(df,aes(percent,fill=perc.type))+geom_histogram(binwidth=1,alpha=.3) +
  facet_wrap(~spp_group) +
  scale_fill_discrete(name="Legend",labels=c("Overlap", "AquaMaps only", "IUCN only"))+
  labs(x='Percent of total range', y='Number of Species')

@@

ggplot(d, aes(sm_perc, fill = sm_range)) + 
  geom_histogram(binwidth = 1, alpha = .7) +
  scale_fill_discrete(name = "Dataset with smaller range", labels = c("AquaMaps", "IUCN")) + 
  labs(x = 'Percent of smaller range that falls within the larger range', y = 'Number of Species')

@@

Exploring possible correlation between area and percent coverage:
``` {r pct overlap vs area, echo = FALSE}

reg <- ggplot(spp_list, aes(total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  geom_abline() + 
  labs(x = 'Total Area (km2)', y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

@@

lm_eqn <- function(df){
    m  <- lm(total_area ~ sm_perc, spp_list)
    eq <- substitute(italic('Total Area') == a + b %.% italic('Percent')*","~~italic(r)^2~"="~r2, 
          list(a  = format(coef(m)[1], digits = 2), 
               b  = format(coef(m)[2], digits = 2), 
               r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq))             
}

print(p1)
```
***

@@

```{r, warning=F,message=F}

#source mapping aquamaps.R to read in all aquamaps raw data. This takes a while so rendering will take a while. 
source('mapping aquamaps.R')

@@

>>>>

library(ggplot2)
library(maps)
library(readr)
library(raster)
library(data.table)
library(tidyr)
library(dplyr)

@@

   p1 = perc_c %>% dplyr::select(percent)

  paste0(round(p1[1, ], 2), '% of the total area in this map is covered by both IUCN and AquaMaps')
  paste0(round(p1[2, ], 2), '% of the total area in this map is covered just by AquaMaps')
  paste0(round(p1[3, ], 2), '% of the total area in this map is covered just by IUCN')
```

@@

This plot shows that the majority of species range maps have only ~30% of their full map covered by both maps, and very few species have more than 50% of their two ranges (IUCN and AquaMaps) overlapping.

@@

overlap = df%>%dplyr::filter(perc.type=='perc')

all<-ggplot(overlap,aes(percent))+
    geom_histogram(aes(y=..count../1926),binwidth=5,alpha=.8)+
    labs(x='Percent of total range that is shared between both datasets', y='Proportion of Total Species')

@@

sp <- ggplot(overlap,aes(percent))+geom_histogram(binwidth=5,alpha=.8) +
  facet_wrap(~spp_group) +
  labs(x='Percent of total range that is shared between both datasets', y='Number of Total Species')

@@

sm_plot <- ggplot(d, aes(sm_perc, fill = sm_range)) + 
       geom_histogram(aes(y=..count../sum(..count..)),binwidth = 4, alpha = .8) +
        scale_fill_discrete(name = "Dataset with smaller range", labels = c("AquaMaps", "IUCN")) + 
        labs(x = 'Percent of smaller range that falls within the larger range', y = 'Proportion of all species (n=1962)')

print(sm_plot)

@@

Exploring possible correlation between total species range area and percent overlap of smaller range within the larger:
``` {r pct overlap vs area, echo = FALSE}

reg <- ggplot(spp_list, aes(sm_perc,total_area, color = sm_range)) + 
  geom_point() + 
  geom_abline() + 
  labs(y = 'Total Area (km2)', x = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

lm_eqn <- function(df){
    m  <- lm(total_area~sm_perc, spp_list)
    eq <- substitute(italic('y') == a + b %.% italic('x')*","~~italic(r)^2~"="~r2, 
          list(a  = format(coef(m)[1], digits = 2), 
               b  = format(coef(m)[2], digits = 2), 
               r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq))             
}

print(p1)

summary(lm(total_area~sm_perc,spp_list))
```

@@

@@

```{r aquamaps probabilities, warning=F,message=F,cache.lazy=FALSE,cache=T,verbose=F,results=F,echo=F}

# get spp_cells raw data for aquamaps

dir_N = c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
          'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

dir_aquamaps = file.path(dir_N,"git-annex/globalprep/SPP_ICO/raw/aquamaps_2014")

spp_cells = fread(input=file.path(dir_aquamaps,"tables/ohi_hcaf_species_native.csv"),verbose=F,header=T,
                  colClasses=c('NULL','character',NA,'numeric','NULL','NULL')) #1 min 49 sec

setnames(spp_cells,c('SpeciesID','CsquareCode','probability'),c("species_id","csquare_code","probability"))

spp_cells = as.data.frame(spp_cells)

spp       = fread(input=file.path(dir_aquamaps,"tables/ohi_speciesoccursum.csv"),verbose=F,header=T)%>%
              mutate(scientific = paste(Genus,Species,sep=" "))
spp = as.data.frame(spp)

@@

```{r,warning=F,message=F, echo=F}

#create dataframe that gives me all species, their total number of cells, total number of cells >=0.4 prob, differnece between the two and threat
df = spp_cells%>%
      group_by(species_id)%>%
        summarize(cells_0.4 = sum(probability>=0.4),
                  cells_t = sum(probability>=0))%>%
          mutate(prop = cells_0.4/cells_t)%>%
            left_join(spp,by=c('species_id'='SPECIESID'))%>%
          select(scientific,species_id,cells_0.4,cells_t,prop,iucn_code)

hist <- ggplot(df,aes(prop,fill=iucn_code))+geom_histogram(aes(y=..count../sum(..count..)),binwidth = 0.05, alpha = .7)+labs(x='Proportion of range area included when selecting a threshold of 0.4',y='Proportion of all aquamaps species (17,348 total)')
print(hist)
